Tropical mathematics and the Lambda-calculus I

### 2023-07-30 CSL24/section5bis.tex
Commit
Dettagli vari

\subparagraph*{Metric approach: bounded $\lambda$-terms as Lipschitz functions}

@@

\subparagraph*{Resource approach: differential $\lambda$-terms as polynomials}

A different family of approaches to linearity and duplication arises from the study of the \emph{differential $\lambda$-calculus} \cite{difflambda} (and differential linear logic \cite{dill}) and its categorical models. 
The key ingredient is a \emph{differential operator} $\Der$,  added to the usual syntax of the $\lambda$-calculus. The intuition is that, given $M$ of type $A\to B$ and $N$ of type $A$, the program $\Der[M]\cdot N$, still of type $A\to B$, corresponds to the \emph{linear application} of $M$ to $N$: this means that $N$ is passed to $M$ so that the latter may use it exactly once (this is why $\Der[M]\cdot N$ still has type $A\to B$, since $M$ might need \emph{other} copies of an input of type $A$).

%Interestingly, the categorical study of the operator $\Der$ has led to the introduction of \emph{cartesian differential categories} $C\partial C$ \cite{Blute2009}, a class of categories providing an abstract axiomatization of differentiation, in the usual mathematical sense. More precisely, a cartesian category $\C C$ is a $C\partial C$ when:
%\begin{itemize}
%\item $\C C$ is left-additive, i.e.~its hom-sets have the structure of commutative monoids, and the cartesian structure is well-behaved w.r.t.~this monoid structure;
%\item $\C C$ is equipped with a differential operator $D:
%\C C(X,Y)\to \C C(X\times X,Y)$ satisfying some axioms which capture usual properties of differentials (e.g.~the linearity of $D$ in one of its two variables, the chain rule, etc.).
%\end{itemize}
%
%Correspondingly, the syntax of the simply typed \emph{differential} $\lambda$-calculus ($\STDLC$) is defined by enriching $\STLC$ with a monoid structure $0,+$ over terms, as well as with $\Der$ and a notion of \emph{linear substitution} (see \cite{difflambda} or the Appendix for details).
%The models of $\STDLC$ are the 
%cartesian \emph{closed} differential categories ($CC\partial C$), 
%which are defined as $C\partial C$ which are also cartesian closed, and in which the monoid structure and the differential operator are both well-behaved with respect to the closed structure \cite{Manzo2012}.

An intriguing similarity between program derivatives and 
actual derivatives is provided by the \emph{Taylor expansion} $\C T$:
in $\STDLC$ one can expand any application $MN$ as an infinite formal sum of \emph{linear} applications
$\Der^{(k)}[M]\cdot N^{k}$, i.e.~where $N$ is passed exactly $k$ times to $M$:
\begin{align}\label{eq:taylor}
\Te{MN} :=  \sum_{k=0}^{\infty}\frac{1}{!k}\cdot (\Der^{(k)}[M] \cdot N^{k})0
\end{align}
In other words, unbounded duplications correspond to some sort of limit of bounded, but arbitrarily large, ones.% (this perspective is made clearer by the related approach of the \emph{resource $\lambda$-calculus} \cite{}).

>>>>

\subparagraph*{Metric approach: bounded $\lambda$-calculus% as Lipschitz functions
}

@@

\subparagraph*{Resource approach: the differential $\lambda$-calculus% as polynomials
}

A different family of approaches to linearity and duplication arises from the study of the \emph{differential $\lambda$-calculus} \cite{difflambda} (and differential linear logic \cite{dill}) and its categorical models. 
The key ingredient is a \emph{differential constructor} $\Der[\_,\_]$,  added to the usual syntax of the $\lambda$-calculus. The intuition is that, given $M$ of type $A\to B$ and $N$ of type $A$, the program $\Der[M,N]$, still of type $A\to B$, corresponds to the \emph{linear application} of $M$ to $N$: this means that $N$ is passed to $M$ so that the latter may use it exactly once (this is why $\Der[M,N]$ still has type $A\to B$, since $M$ might need \emph{other} copies of an input of type $A$). 
%Interestingly, the categorical study of the operator $\Der$ has led to the introduction of \emph{cartesian differential categories} $C\partial C$ \cite{Blute2009}, a class of categories providing an abstract axiomatization of differentiation, in the usual mathematical sense. More precisely, a cartesian category $\C C$ is a $C\partial C$ when:
%\begin{itemize}
%\item $\C C$ is left-additive, i.e.~its hom-sets have the structure of commutative monoids, and the cartesian structure is well-behaved w.r.t.~this monoid structure;
%\item $\C C$ is equipped with a differential operator $D:
%\C C(X,Y)\to \C C(X\times X,Y)$ satisfying some axioms which capture usual properties of differentials (e.g.~the linearity of $D$ in one of its two variables, the chain rule, etc.).
%\end{itemize}
%
%Correspondingly, the syntax of the simply typed \emph{differential} $\lambda$-calculus ($\STDLC$) is defined by enriching $\STLC$ with a monoid structure $0,+$ over terms, as well as with $\Der$ and a notion of \emph{linear substitution} (see \cite{difflambda} or the Appendix for details).
%The models of $\STDLC$ are the 
%cartesian \emph{closed} differential categories ($CC\partial C$), 
%which are defined as $C\partial C$ which are also cartesian closed, and in which the monoid structure and the differential operator are both well-behaved with respect to the closed structure \cite{Manzo2012}. 
The reason why $\Der$ is called a ``differential'', is twofold: semantically, its interpretation is a generalisation of the usual differential (see \autoref{section3}); syntactically, it allows to define the so-called \emph{Taylor expansion} $\C T$ of programs:
the idea is that, in $\STDLC$, one can expand any application $MN$ as an infinite formal sum of \emph{linear} applications
$\Der^{k}[M,N^k]$, i.e.~where $N$ is linearly passed exactly $k$ times to $M$, and doing this recursively gives rise to the series $\Te{MN} :=  \sum_{k=0}^{\infty}\frac{1}{!k}\cdot \Der^{k}[\Te{M},\Te{N}^k]0$.
In other words, unbounded duplications correspond to some sort of limit of bounded, but arbitrarily large, ones.
Actually, we will work with the so-called \emph{qualitative} Taylor expansion, where one quotients for idempotency of the $+$ operator in the series, and thus the Taylor expansion is just a set (see \autoref{sec:TayLip}).% (this perspective is made clearer by the related approach of the \emph{resource $\lambda$-calculus} \cite{}).

### 2023-07-30 CSL24/sectionTaylip.tex
Commit
Dettagli vari

The key ingredient is the notion of Taylor expansion of $\lambda$-terms, that we now quickly recall. 
Let $M\cdot \langle N_{1},\dots, N_{k}\rangle$ be an abbreviation for
$\mathsf D^{n}[M]\cdot \langle N_{1},\dots, N_{k}\rangle$. A \emph{resource $\lambda$-term} is any term belonging to the grammar $t::= x \mid \lambda x.t \mid t\cdot \langle t,\dots, t\rangle$.
Observe that in a resource term applications are always \emph{bounded}: they may use an exact number of copies of their input.

The Taylor expansion of a $\lambda$-term is a set of resource terms $\Te{M}$ defined inductively as $\Te{x}=\{x\}$, $\Te{\lambda x.M}=\{\lambda x.t\mid t\in \Te{M}\}$ and 
$\Te{MN}=\{ t\cdot \langle u_{1},\dots, u_{k}\rangle \mid k\in \mathbb N, t\in \Te{M}, u_{i}\in \Te{N} \}$. 
Considering the term $M=zxx$ from Example \ref{ex:zxx}, all resource terms
 $t_{n,m}=(z\cdot x^{n})\cdot x^{m}$, for $n,m\in \mathbb N$, are in $\Te{M}$. 
Observe that the interpretation of $t_{n,m}$ yields a tropical polynomial
$\model{t_{n,m}}^{!}(x)(z)= y_{[n,m]}+(n+m)x$, rather than a tLs. 
However, resource term need not always yield a tropical polynomial. For instance, consider $y:(*\to*)\to (*\to*), x:(*\to *) \vdash t:  (*\to  *)$
with $t=y\cdot (y\cdot x)$. Then 
$\model{t}^{!}: \Lawv^{  \mathcal M_{\mathrm{fin}}(\mathbb N)\times \mathbb N}
\times \Lawv^{\mathbb N} \to \Lawv^{\mathbb N}$ is given by
$
\model{t}^{!}(y,x)_{i}= \inf_{m,n\in \mathbb N}\big\{    y_{[m],i}  +  y_{[n],m}+  x_{n}\}
$, 
which is not a polynomial. Yet, $\model{t}^{!}$ is Lipschitz, more precisely, 1-Lipschitz in $x$ and 2-Lipschitz in $y$. This is a general fact, as shown below.

@@

\begin{theorem}
Let $\mathcal S\in \{\RM{PCF}^{\Lawv}, \STLC,\BSTLC,\STDLC\}$, $\Gamma\vdash_{\mathcal S}M:A$; for all $a\in \model{A}$,
\begin{enumerate}
\item if $\mathcal S=\BSTLC$, then $\model{\Gamma\vdash_{\mathcal S}M:A}^{!}_{a}$ is a tropical polynomial, and thus Lipschitz;

\item if $\mathcal S=\STDLC$ and $M$ is a resource term, then $\model{\Gamma\vdash_{\mathcal S}M:A}^{!}_{a}$ is Lipschitz;

\item if $\mathcal S=\STLC,\RM{PCF}^{\Lawv}$, then $\model{\Gamma\vdash_{\mathcal S}M:A}^{!}_{a}$ is locally Lipschitz;

\item if $\mathcal S=\STLC$, then $\Te{M}$ decomposes $\model{\Gamma \vdash_{\STLC} M:A}^!_{a}$ as an $\inf_{t\in\Te{M}}\model{\Gamma\vdash_{\STDLC} t:A}^!_{a}$ of {Lipschitz} functions.
\end{enumerate}

@@

\begin{example}
Consider again the term $M=zxx$ from Example \ref{ex:zxx}-
%, satisfying $x:*, z:(*\to*\to*) \vdash M:  *$. Since $\mathcal M_{\mathrm{fin}}(\{*\})\simeq \mathbb N$, we have that 
%
% $\model{M}^{!}: \Lawv \times \Lawv^{\mathcal M_{\mathrm{fin}}(\mathbb N\times \mathbb N)}\to \Lawv$ is given by
 The (generalized) tLs $\model{M}^{!}(x)(y)= \inf_{n,n'\in \mathbb N}\{y_{[(n,n')]}+(n+n')x\}$ is not (globally) Lipschitz: for any 
$L>0$, choose a natural number $N>L$, let $Y\in \Lawv^{\mathcal M_{\mathrm{fin}}(\mathbb N\times \mathbb N)}$ be such that $Y_{\mu}<\infty$ only if $\mu=[(n,n')]$ with $n+n'\geq N$; then $|\model{M}^{!}(x)(Y)- \model{M}^{!}(x+\epsilon)(Y)|\geq N\epsilon > L\epsilon$. 
Now take the approximant $t= (z\cdot \langle x^{N-1}\rangle) \cdot \langle x\rangle \in \Te{M}$ (chosen so that $\model t^{!}(x)(Y)<\infty$). Its interpretation is the monomial 
$\model{t}^{!}(x)(Y) = Y_{[(N-1,1)]}+Nx$. We can then compute a Lipschitz-constant for $\model{M}^{!}$ around $\langle x,Y\rangle$
as $\frac{1}{\delta}\model{t}^{!}(\langle x,Y\rangle+\delta)= N+2 + \frac{Y_{[(N-1,1)]}+Nx}{\delta}$. 
\end{example}
%    
%    %
%  Consider now $F\in \Lawv^{\mathcal M_{\mathrm{fin}}(\mathbb N\times \mathbb N)}$ given by $F_{\mu}=0$ if $\mu=[(1,1)]$ and $F_{\mu}=\infty$ otherwise. 
%  Then $\model{M}^{!}(\langle x,F\rangle)=2x$
%%We show that we can compute a local 
%More formally, the differential operator $\Der[-]$ transforms a function $M:A\to B$ into a function $\Der[M]: A\to (A\to B)$ which is linear in its first argument. 
%Since $M$ may rather ask for several copies of $N$, this requires a form of non-determinism: 
%For example, if $M$ is the term $\lambda fx.f(fx)$ considered before, $\Der[M]$ takes a first input $N$ and passes it linearly to $M$. Notice that there are two ways of doing so, corresponding to the two bound occurrences of $f$ in $M$: either by applying $N$ to $fx$, or by 
%applying $f$ linearly to $Nx$ (indeed, if $f$ were applied in an unrestricted way, it might duplicate $Nx$, so that $N$ would not be used linearly). This justifies the equation below, in which $\Der[M]$ is identified with the non-deterministic sum of the two possible linear choices:
%\begin{align}
%\Der\left[\lambda f x.f(fx)\right]\cdot N = 
%\lambda fx. N(fx) + \left(\Der[f]\cdot (Nx)\right)(fx)
%\end{align}
%More generally, one can define a notion of $k$-bounded application $\Der^{(k)}[M]\cdot N^{k}$, where $\Der^{(0)}[M]\cdot N^{0}= M$ and $\Der^{(k+1)}[M]\cdot N^{k+1}= \Der[ \Der^{(k)}[M]\cdot N^{k}]\cdot N$, corresponding to passing $N$ to $M$ exactly $k$ times.
%
%
%The name ``differential'' for the operator $\Der[-]$ is justified by the fact that it satisfies many properties of the usual differential operator of analysis $\Der[f]:= \lambda xy. \frac{\mathsf df(y)}{\mathsf dy}\cdot x$. Notably, it is additive in its first variable (i.e.~it commutes with the non-deterministic sum operator), and satisfies the chain rule.
%Most famously, the differential operator can be used to define a Taylor formula for $\lambda$-terms, which decomposes an unrestricted application into a formal non-deterministic sum of bounded applications:

>>>>

The key ingredient is the notion of Taylor expansion of $\lambda$-terms, that we now quickly recall. 
Let $M\langle N_{1},\dots, N_{k}\rangle$ be an abbreviation for
$\mathsf D^{k}[M, N_{1},\dots, N_{k}]0$. A \emph{resource $\lambda$-term} is any term belonging to the grammar $t::= x \mid \lambda x.t \mid t \langle t,\dots, t\rangle$.
Observe that in a resource term applications are always \emph{bounded}: they may use an exact number of copies of their input.

The Taylor expansion of a $\lambda$-term is a set of resource terms $\Te{M}$ defined inductively as $\Te{x}=\{x\}$, $\Te{\lambda x.M}=\{\lambda x.t\mid t\in \Te{M}\}$ and 
$\Te{MN}=\{ t \langle u_{1},\dots, u_{k}\rangle \mid k\in \mathbb N, t\in \Te{M}, u_{i}\in \Te{N} \}$. 
Considering the term $M=zxx$ from Example \ref{ex:zxx}, all resource terms
 $t_{n,m}=z \,\langle x^{n}\rangle\,\langle x^{m}\rangle$, for $n,m\in \mathbb N$, are in $\Te{M}$. 
Observe that the interpretation of $t_{n,m}$ yields a tropical polynomial
$\model{t_{n,m}}^{!}(x)(z)= y_{[n,m]}+(n+m)x$, rather than a tLs. 
However, resource term need not always yield a tropical polynomial. For instance, consider $y:(*\to*)\to (*\to*), x:(*\to *) \vdash t:  (*\to  *)$
with $t=y\cdot (y\cdot x)$. Then 
$\model{t}^{!}: \Lawv^{  \mathcal M_{\mathrm{fin}}(\mathbb N)\times \mathbb N}
\times \Lawv^{\mathbb N} \to \Lawv^{\mathbb N}$ is given by
$
\model{t}^{!}(y,x)_{i}= \inf_{m,n\in \mathbb N}\big\{    y_{[m],i}  +  y_{[n],m}+  x_{n}\}
$, 
which is not a polynomial. Yet, $\model{t}^{!}$ is Lipschitz, more precisely, 1-Lipschitz in $x$ and 2-Lipschitz in $y$. This is a general fact, as shown below.

@@

\begin{theorem}
Let $\mathcal S$ be one of $\RM{PCF}^{\Lawv},\STLC,\BSTLC,\STDLC$. Let $\Gamma\vdash_{\mathcal S}M:A$ and $a\in \model{A}$.
\begin{enumerate}
\item For $\mathcal S=\BSTLC$, $\model{\Gamma\vdash_{\mathcal S}M:A}^{!}_{a}$ is a tropical polynomial, and thus Lipschitz;

\item For $\mathcal S=\STDLC$, if $t$ is a resource term, then $\model{\Gamma\vdash_{\mathcal S}t:A}^{!}_{a}$ is Lipschitz;

\item For $\mathcal S=\STLC,\RM{PCF}^{\Lawv}$, then $\model{\Gamma\vdash_{\mathcal S}M:A}^{!}_{a}$ is locally Lipschitz;

\item For $\mathcal S=\STLC$, $\Te{M}$ decomposes $\model{\Gamma \vdash_{\STLC} M:A}^!_{a}$ as an $\inf_{t\in\Te{M}}\model{\Gamma\vdash_{\STDLC} t:A}^!_{a}$ of {Lipschitz} functions.
\end{enumerate}

@@

\begin{example}
Consider again the term $M=zxx$ from Example \ref{ex:zxx}-
%, satisfying $x:*, z:(*\to*\to*) \vdash M:  *$. Since $\mathcal M_{\mathrm{fin}}(\{*\})\simeq \mathbb N$, we have that 
%
% $\model{M}^{!}: \Lawv \times \Lawv^{\mathcal M_{\mathrm{fin}}(\mathbb N\times \mathbb N)}\to \Lawv$ is given by
 The (generalized) tLs $\model{M}^{!}(x)(y)= \inf_{n,n'\in \mathbb N}\{y_{[(n,n')]}+(n+n')x\}$ is not (globally) Lipschitz: for any 
$L>0$, choose a natural number $N>L$, let $Y\in \Lawv^{\mathcal M_{\mathrm{fin}}(\mathbb N\times \mathbb N)}$ be such that $Y_{\mu}<\infty$ only if $\mu=[(n,n')]$ with $n+n'\geq N$; then $|\model{M}^{!}(x)(Y)- \model{M}^{!}(x+\epsilon)(Y)|\geq N\epsilon > L\epsilon$. 
Now take the approximant $t= z \langle x^{N-1}\rangle \langle x\rangle \in \Te{M}$ (chosen so that $\model t^{!}(x)(Y)<\infty$). Its interpretation is the monomial 
$\model{t}^{!}(x)(Y) = Y_{[(N-1,1)]}+Nx$. We can then compute a Lipschitz-constant for $\model{M}^{!}$ around $\langle x,Y\rangle$
as $\frac{1}{\delta}\model{t}^{!}(\langle x,Y\rangle+\delta)= N+2 + \frac{Y_{[(N-1,1)]}+Nx}{\delta}$. 
\end{example}
%    
%    %
%  Consider now $F\in \Lawv^{\mathcal M_{\mathrm{fin}}(\mathbb N\times \mathbb N)}$ given by $F_{\mu}=0$ if $\mu=[(1,1)]$ and $F_{\mu}=\infty$ otherwise. 
%  Then $\model{M}^{!}(\langle x,F\rangle)=2x$
%%We show that we can compute a local 
%More formally, the differential operator $\Der[-]$ transforms a function $M:A\to B$ into a function $\Der[M]: A\to (A\to B)$ which is linear in its first argument. 
%Since $M$ may rather ask for several copies of $N$, this requires a form of non-determinism: 
%For example, if $M$ is the term $\lambda fx.f(fx)$ considered before, $\Der[M]$ takes a first input $N$ and passes it linearly to $M$. Notice that there are two ways of doing so, corresponding to the two bound occurrences of $f$ in $M$: either by applying $N$ to $fx$, or by 
%applying $f$ linearly to $Nx$ (indeed, if $f$ were applied in an unrestricted way, it might duplicate $Nx$, so that $N$ would not be used linearly). This justifies the equation below, in which $\Der[M]$ is identified with the non-deterministic sum of the two possible linear choices:
%\begin{align}
%\Der\left[\lambda f x.f(fx)\right]\cdot N = 
%\lambda fx. N(fx) + \left(\Der[f]\cdot (Nx)\right)(fx)
%\end{align}
%More generally, one can define a notion of $k$-bounded application $\Der^{(k)}[M]\cdot N^{k}$, where $\Der^{(0)}[M]\cdot N^{0}= M$ and $\Der^{(k+1)}[M]\cdot N^{k+1}= \Der[ \Der^{(k)}[M]\cdot N^{k}]\cdot N$, corresponding to passing $N$ to $M$ exactly $k$ times.
%
%
%The name ``differential'' for the operator $\Der[-]$ is justified by the fact that it satisfies many properties of the usual differential operator of analysis $\Der[f]:= \lambda xy. \frac{\mathsf df(y)}{\mathsf dy}\cdot x$. Notably, it is additive in its first variable (i.e.~it commutes with the non-deterministic sum operator), and satisfies the chain rule.
%Most famously, the differential operator can be used to define a Taylor formula for $\lambda$-terms, which decomposes an unrestricted application into a formal non-deterministic sum of bounded applications:
